Marginally-Stable Thermal Equilibria of Rayleigh-Bénard Convection

### 2021-04-15 publication_materials/ms_rbc.tex
pre touch up

\begin{abstract}
Natural convection is ubiquitous throughout the physical sciences and engineering, yet many of its important properties remain elusive -- particularly in the large Rayleigh number regime.
In this investigation, we derive and solve a quasilinear form of the Boussinesq Rayleigh-B\'enard equations. 
Eigenfunction amplitudes are obtained by requiring marginal-stability. We obtain symmetric thermally-equilibrated solutions and analyze their properties. 
Solutions to the quasilinear problem differ from statistically-steady 2-dimensional Rayleigh-Be\'nard convection simulation data, having thinner boundary layers and larger Nusselt numbers $\Nu$ which scale like $\Nu \sim Ra^{1/3}$. 
We initialize 2-dimensional simulations with solutions to the quasilinear problem; they do not undergo a convective transient period and are kinetically exaggerated.
\end{abstract}

@@

\begin{figure}
    \includegraphics[width=3.4in]{EV_spectrum_ol.png}
    \caption{Eigenvalue spectra for $Ra = 10^5$. The spectrum of an ``initial" marginally-stable mean temperature profile $\bar{T}(z, t_0)$ has a maximum eigenvalue of 0. 
    Given a small fixed timestep $\Delta t$, diffusion destabilizes the system, increasing its eigenvalues. 
    Advection tends to stabilize the system, decreasing its eigenvalues. 
    It follows that there exists some proportion ($A^2(t_0)$) of these two terms which yields a new marginally-stable mean temperature profile $\bar{T}(z, t_0 + \Delta t)$, whose spectrum is ``evolved".}
    \label{fig:iteration_spectra} 
\end{figure}

@@

\subsection{Treatment of multiple marginally-stable modes} \label{sec:multiple_modes}
In most cases, we eventually encounter eigenvalue spectra with multiple simultaneously marginal modes. If ignored, we are forced to reduce the timestep and allow the modes to alternate. Instead, we generalize the advective term in (\ref{EQ:T0_IVP}) to accommodate $N$ simultaneously marginal modes
\begin{equation}
    A^2 \langle w' b' \rangle_x = \sum_{n = 1}^{N} A^2_{n} \langle w' b' \rangle_{x, \, n}
\end{equation}
where  $\langle w' b' \rangle_{x, \, n}$ is composed of perturbations associated with $k_x = \frac{n\pi}{2}$. 
There are now $N$ amplitudes to solve for and $N$ eigenvalues to keep marginally-stable. 
In general, we expect a function $\mathbf{\omega} \, : \, \mathbb{R}^N \to  \mathbb{R}^N$ to have isolated roots (should they exist). 
We employ Broyden's method for root-finding in multi-dimensional functions whose derivatives are not explicitly known. 
Difficulty arises when transitioning between different numbers of marginal modes. 
Here we rely on $A^2 > 0$ by asserting that a mode which is \textit{close enough} to marginal-stability can be included in the iteration provided that its respective amplitude is positive. 
Should we converge upon a negative amplitude, that mode is discarded and the iteration is repeated. 
In general we find that no other obvious course of action yields equilibria.

@@

The most resilient and unexpected feature of thermally equilibrated mean temperature profiles are the pronounced dips adjacent to the boundary layers. 
These dips appear in every solution, regardless of $Ra$. 
Physically, they correspond to thin layers in which the temperature gradient reverses, contradicting an important hypothesis of \cite{Malkus_1954}. 
This counter-diffusion, which opposes overall heat transfer, is overcome by the coinciding advective flux, shown in Figure \ref{fig:flux}.  
\begin{figure}
    \centering
    \includegraphics[width=3.4in]{T_profs_na.png}
    \caption{Mean temperature profiles $\bar{T}$ for $Ra = 10^8$. 
    The initial profile is given by (\ref{EQ:T0}). 
    The MSTE curve refers to the mean temperature profile of the marginally-stable thermally-equilibrated state. 
    The DNS profile is obtained by performing a 2-dimensional nonlinear simulation of (\ref{EQ:motion1})--(\ref{EQ:motion3}) with \texttt{Dedalus} until thermal and kinetic relaxation. 
    The DNS temperature data are horizontally- and time-averaged. 
    The initial, MSTE, and DNS profiles have increasingly relaxed boundary regions (respectively). 
    The MSTE profile exhibits prominent dips, nested alongside the boundary regions. 
    The source of this feature is not well understood, but similar temperature gradient reversal regions were found by \cite{chini_cells} along the midplanes of 2-dimensional convective cellular solutions at $Ra \sim 10^6$.}
    \label{fig:T0_profiles}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=3.4in]{kx_m_ra1.png}
    \caption{Wavenumbers of marginally-stable modes of thermally equilibrated states. 
    Markers are color-coded according to their adjacent local maxima index in the eigenvalue spectrum. 
    For example, the spectrum corresponding to $Ra = 10^5$ has adjacent marginal wavenumbers $k_x = \pi, \, 1.5\pi$. 
    The $Ra = 10^9$ spectrum, shown in lower right corner of Figure \ref{fig:flux}, has three groups of maxima, with a single marginal mode in the first group ($k_x = 1.5\pi$), two adjacent marginal modes in the second group ($k_x = 6\pi, \, 6.5\pi$), and two adjacent marginal modes in the third group ($k_x = 23.5\pi, \, 24\pi$). 
    The largest wavenumbers of the green branch obey a power-law relationship with $Ra$}
    \label{fig:kx_marginals}
\end{figure}
\begin{figure*}
    \centering
    \begin{tabular}{@{}c@{}}
        \includegraphics[width=3.4in]{flux_sup_n.png}
    \end{tabular}
    \begin{tabular}{@{}c@{}}
        \includegraphics[width=3.4in]{EV_spectra_2ra.png}
    \end{tabular}
    \caption{Heat fluxes (left) and eigenvalue spectra (right) of equilibrated states $Ra = 2 \times 10^5$ (top) and $Ra = 10^9$ (bottom). 
    Advection profiles belong to marginally-stable modes. 
    For low $Ra$, a single mode with $k_x = 1.5\pi$ is sufficient to oppose boundary layer diffusion and facilitate heat flux throughout the bulk of the domain. 
    For large $Ra$, high-wavenumber modes contribute pronounced small-scale advection profiles which tightly hug the thin boundary layers. 
    A combination of of progressively wider advection profiles is necessary to transition to the $k_x = 1.5\pi$ mode.}
    \label{fig:flux}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=7.1in]{nu_ra.PNG}
    \caption{Nusselt numbers are shown for marginally-stable thermally-equilibrated states, statistically-steady direct numerical simulations performed with \texttt{Dedalus} \cite{Anders_cd}, and equilibrated nonlinear solutions, which maximize heat flux, given by \cite{Waleffe}. 
    Both obey power-law relationships with $Ra$, with $\Nu \sim Ra^{1/3}$ in agreement with Figures \ref{fig:bl_ra}, \ref{fig:kx_marginals}, and \ref{fig:del_inv} as well as \cite{Malkus_1954}. 
    Marginally-stable states' Nusselt numbers exceed simulation Nusselt numbers. 
    This can be explained by the contrasting boundary region geometries shown in Figure \ref{fig:T0_profiles}.}%
    \label{fig:nu_vs_ra}%
\end{figure*}

@@

The Nusselt number, which describes convective performance, is given by
\begin{equation}
  \Nu = \frac{\langle A^2 \langle w'T' \rangle_x - \mathcal{P}\frac{\partial \bar{T}}{\partial z} \rangle_z}{\langle- \mathcal{P}\frac{\partial \bar{T}}{\partial z} \rangle_z}.
\end{equation}
There is no general consensus surrounding the scaling behavior $\Nu \sim Ra^{\beta}$ for high $Ra$ systems. The asymptotic value of $\beta$ is of particular importance in astrophysical systems. In Figure \ref{fig:nu_vs_ra} we report $\Nu$ for MSTE, DNS, along with several ECS. 
The marginally-stable thermally-equilibrated states and numerical simulation data exhibit power-law relationships with $Ra$. 
More specifically, $\beta = 1/3$ which agrees with Figures \ref{fig:bl_ra}, \ref{fig:kx_marginals}, and \ref{fig:del_inv} and the scaling argument given by \cite{Malkus_1954}. 
The nonlinear solutions, which maximize heat flux, are obtained by \cite{Waleffe}. 
$\Nu$ of both equilibria exceed that of simulations, which might be due to the transitional behaviors outlined by \cite{Yalniz} and \cite{Cvitanovic} inhibiting heat flux. 
Alternatively, we might anticipate the existence of exact coherent states with smaller $\Nu$, forming another node in the Markov chain whose long-term behavior agrees with the simulation data.

@@

% \begin{figure}
%     \centering
%     \includegraphics[width=3.4in]{sim_eq_nu.png}
%     \caption{Nusselt numbers of simulations performed at $Ra \, = \, 10^8$ with conventional initial conditions (red) and marginally-stable thermally-equilibrated initial conditions (blue). Simulations launched with thermally-equilibrated states do not undergo a convective-transient period because the characteristic plume structure exists on initialization. This is illustrated by the $Nu$ spike at $t \sim 15$ in the convectional simulation.}
%     \label{fig:nu_sim}
% \end{figure}
The marginally-stable equilibria are associated with a combination of eigenfunctions whose average kinetic energy is significantly larger than the statistical steady state, as given by Figure \ref{fig:ke_sim}. 
This is postpones kinetic relaxation because the large horizontal flows decay on a viscous timescale
\begin{equation}
    t_{\nu} \sim \sqrt{\frac{Ra}{Pr}}. \nonumber
\end{equation}
Consequently, marginally-stable thermally-equilibrated initial conditions do not reduce the simulation time required to achieve a statistically steady state, rather, they increase it considerably. Further, this suggests that the MSTE background state perspective is, at least partially, flawed as an appropriate background state would approximate $\langle|\mathbf{u}|^2 \rangle_{\mathcal{D}}$
% \begin{figure}
%     \centering
%     \includegraphics[width=3.4in]{sim_eq_ke.png}
%     \caption{Average kinetic energies of the same simulations performed in Figure \ref{fig:nu_sim} ($Ra = 10^8$). The combination of eigenfunctions which give thermal equilibrium have significantly more kinetic energy than the statistically steady state. Kinetic equilibrium is achieved according to the viscous timescale $t_{\nu} \sim Pr Ra^{-1/2}$.}
%     \label{fig:ke_sim}
% \end{figure}
\begin{figure}
    \begin{minipage}{3.4in}
        \centering
        \includegraphics[width=3.4in]{sim_eq_nu.png}
        \caption{Nusselt numbers of simulations performed at $Ra = 10^8$ with the conventional initial conditions (black) and MSTE initial conditions (blue). 
        MSTE simulations do not undergo a convective-transient period because the characteristic large-scale convective cell structure exists on initialization. For high $Ra$ simulations, researchers often choose to recycle the results of similar $Ra$ runs \cite{Verzicco, Johnston} because the high-Reynolds number flows associated with the convective transient period are computationally prohibitive \cite{Anders_AE}.}
        \label{fig:nu_sim}
    \end{minipage}
\end{figure}

@@

We cannot use our reduced quasilinear model to predict $\Nu \sim Ra^{\beta}$ for large $Ra$ without some complementary stability constraint. 
The eigenvalues of DNS data might provide an informative stability constraint (we could find MUTE and compare them with DNS data), but this would be self-defeating in general. 
Instead we seek to combine reduced models to approximate $\beta$ without DNS. Under the marginal-stability constraint, we found that MSTE exhibit classical Malkus scaling, having marginally-stable boundary layers with $\beta = 1/3$. 
We could instead require that the eigenvalues of MUTE agree with the eigenvalues of the analytic 1-dimensional boundary layer predictions of \cite{Shishkina, Zhang_20} (which remain in development), allowing us to refine and expand these estimates into 2-dimensional MUTE. 
Accessible MUTE states would not rely on DNS, satisfying only the quasilinear problem and stability constraint. 
Such MUTE ought to agree with DNS data more precisely then MSTE, ideally having less extreme velocities. 
This is reasonable to expect, since accurate boundary layer approximations and DNS data should have similar stability characteristics. 
By definition, $\bar{T}_{\rm{MUTE}}$ is unstable and would therefore admit more diffuse boundary layer structures, requiring smaller $A^2(t)$ (less advection) to satisfy the stability constraint and thereby reducing $\Nu$, $\langle \langle |\mathbf{u}|^2 \rangle_x \rangle_z$ and possibly $\beta$.

>>>>

\begin{abstract}
Natural convection is ubiquitous throughout the physical sciences and engineering, yet many of its important trends remain elusive -- particularly in the large Rayleigh number regime.
In this investigation, we derive and solve a quasilinear form of the Boussinesq Rayleigh-B\'enard equations. 
Eigenfunction amplitudes are obtained by requiring marginal-stability. We obtain symmetric thermally-equilibrated solutions and analyze their properties. 
Solutions to the quasilinear problem differ from statistically-steady 2-dimensional Rayleigh-Be\'nard convection simulation data, having thinner boundary layers and larger Nusselt numbers $\Nu$ which scale like $\Nu \sim Ra^{1/3}$. 
We initialize 2-dimensional simulations with solutions to the quasilinear problem; they do not undergo a convective transient period and are kinetically exaggerated.
\end{abstract}

@@

\begin{figure}
    \includegraphics[width=3.4in]{EV_spectrum_ol.png}
    \caption{Eigenvalue spectra for $Ra = 10^5$. The spectrum of an ``initial" marginally-stable mean temperature profile $\bar{T}(z, t_0)$ has a maximum eigenvalue of 0. 
    Given a small fixed timestep $\Delta t$, diffusion destabilizes the system, increasing its eigenvalues. 
    Advection tends to stabilize the system, decreasing its eigenvalues. 
    It follows that there exists some proportion $A^2(t_0)$ of these two terms which yields a new marginally-stable mean temperature profile $\bar{T}(z, t_0 + \Delta t)$, whose spectrum is ``evolved".}
    \label{fig:iteration_spectra} 
\end{figure}

@@

\subsection{Treatment of multiple marginally-stable modes} \label{sec:multiple_modes}
In most cases, we eventually encounter eigenvalue spectra with multiple simultaneously marginal modes. If ignored, we are forced to reduce the timestep and allow the modes to alternate. Instead, we generalize the advective term in (\ref{EQ:T0_IVP}) to accommodate $N$ simultaneously marginal modes
\begin{equation}
    A^2 \langle w' T' \rangle_x = \sum_{n = 1}^{N} A^2_{n} \langle w' T' \rangle_{x, \, n}
\end{equation}
where  $\langle w' T' \rangle_{x, \, n}$ is composed of perturbations associated with $k_x = \frac{n\pi}{2}$. 
There are now $N$ amplitudes to solve for and $N$ eigenvalues to keep marginally-stable. 
In general, we expect a function $\mathbf{\omega} \, : \, \mathbb{R}^N \to  \mathbb{R}^N$ to have isolated roots (should they exist). 
We employ Broyden's method for root-finding in multi-dimensional functions whose derivatives are not explicitly known. 
Difficulty arises when transitioning between different numbers of marginal modes. 
Here we rely on $A^2 > 0$ by asserting that a mode which is \textit{close enough} to marginal-stability can be included in the iteration provided that its respective amplitude is positive. 
Should we converge upon a negative amplitude, that mode is discarded and the iteration is repeated. 
In general we find that no other obvious course of action yields equilibria.

@@

The most resilient and unexpected feature of thermally equilibrated mean temperature profiles are the pronounced dips adjacent to the boundary layers. 
These dips appear in every solution, regardless of $Ra$. 
Physically, they correspond to thin layers in which the temperature gradient reverses, contradicting an important hypothesis of \cite{Malkus_1954}. 
This counter-diffusion, which opposes overall heat transfer, is overcome by the coinciding advective flux, shown in Figure \ref{fig:flux}.  
\begin{figure}
    \centering
    \includegraphics[width=3.4in]{T_profs_na.png}
    \caption{Mean temperature profiles $\bar{T}$ for $Ra = 10^8$. 
    The initial profile is given by (\ref{EQ:T0}). 
    The MSTE curve refers to the mean temperature profile of the marginally-stable thermally-equilibrated state. 
    The DNS profile is obtained by performing a 2-dimensional nonlinear simulation of (\ref{EQ:motion1})--(\ref{EQ:motion3}) with \texttt{Dedalus} until thermal and kinetic relaxation. 
    The DNS temperature data are horizontally- and time-averaged. 
    The initial, MSTE, and DNS profiles have increasingly relaxed boundary regions (respectively). 
    The MSTE profile exhibits prominent dips, nested alongside the boundary regions. 
    The source of this feature is not well understood, but similar temperature gradient reversal regions were found by \cite{chini_cells} along the midplanes of 2-dimensional convective cellular solutions at $Ra \sim 10^6$.}
    \label{fig:T0_profiles}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=3.4in]{kx_m_ra1.png}
    \caption{Wavenumbers of marginally-stable modes of thermally equilibrated states. 
    Markers are color-coded according to their adjacent local maxima index in the eigenvalue spectrum. 
    For example, the spectrum corresponding to $Ra = 10^5$ has adjacent marginal wavenumbers $k_x = \pi, \, 1.5\pi$. 
    The $Ra = 10^9$ spectrum, shown in lower right corner of Figure \ref{fig:flux}, has three groups of maxima, with a single marginal mode in the first group ($k_x = 1.5\pi$), two adjacent marginal modes in the second group ($k_x = 6\pi, \, 6.5\pi$), and two adjacent marginal modes in the third group ($k_x = 23.5\pi, \, 24\pi$). 
    The largest wavenumbers of the green branch obey a power-law relationship with $Ra$}
    \label{fig:kx_marginals}
\end{figure}
\begin{figure*}
    \centering
    \begin{tabular}{@{}c@{}}
        \includegraphics[width=3.4in]{flux_sup_n.png}
    \end{tabular}
    \begin{tabular}{@{}c@{}}
        \includegraphics[width=3.4in]{EV_spectra_2ra.png}
    \end{tabular}
    \caption{Heat fluxes (left) and eigenvalue spectra (right) of equilibrated states $Ra = 2 \times 10^5$ (top) and $Ra = 10^9$ (bottom). 
    Advection profiles belong to marginally-stable modes. 
    For low $Ra$, a single mode with $k_x = 1.5\pi$ is sufficient to oppose boundary layer diffusion and facilitate heat flux throughout the bulk of the domain. 
    For large $Ra$, high-wavenumber modes contribute pronounced small-scale advection profiles which tightly hug the thin boundary layers. 
    A combination of of progressively wider advection profiles is necessary to transition to the $k_x = 1.5\pi$ mode.}
    \label{fig:flux}
\end{figure*}
\begin{figure*}
    \centering
    \includegraphics[width=7.1in]{nu_ra.PNG}
    \caption{Nusselt numbers are shown for MSTE, aspect-ratio optimized ECS \cite{Wen}, and statistically-steady DNS \cite{Johnston, Scheel_2016}.
    All datasets obey power-law relationships, with MSTE admitting classical scaling $\Nu \sim Ra^{1/3}$. 
    MSTE facilitate more heat transfer than the ECS and DNS, as can be explained by the contrasting boundary region geometries shown in Figure \ref{fig:T0_profiles}.}%
    \label{fig:nu_vs_ra}%
\end{figure*}

@@

The Nusselt number, which measures convective performance, is given by
\begin{equation}
  \Nu = \frac{\langle A^2 \langle w'T' \rangle_x - \mathcal{P}\frac{\partial \bar{T}}{\partial z} \rangle_z}{\langle- \mathcal{P}\frac{\partial \bar{T}}{\partial z} \rangle_z}.
\end{equation}
There is no general consensus surrounding the scaling behavior of $\Nu$ for high $Ra$ systems, which are of particular importance in astrophysical systems. In Figure \ref{fig:nu_vs_ra} we report $\Nu$ for MSTE and DNS states. 
Both exhibit power-law dependence $\Nu \sim Ra^{\beta}$ with classical Malkus scaling $\beta = 1/3$ \cite{Malkus_1954}. 
\cite{Wen} hypothesized that $\Nu$ of ECS which admit classical Malkus scaling must always exceed $\Nu$ of turbulent convection.
Our findings agree; $\Nu$ of MSTE exceed 2D and 3D DNS. 
This might be due to the chaotic transitions among unstable periodic orbits outlined by \cite{Yalniz, Cvitanovic} inhibiting heat flux. 
We might also anticipate the existence of exact coherent states with smaller $\Nu$, occupying neighboring nodes in the Markov chain whose long-term behavior agrees with DNS.

@@

% \begin{figure}
%     \centering
%     \includegraphics[width=3.4in]{sim_eq_nu.png}
%     \caption{Nusselt numbers of simulations performed at $Ra \, = \, 10^8$ with conventional initial conditions (red) and marginally-stable thermally-equilibrated initial conditions (blue). Simulations launched with thermally-equilibrated states do not undergo a convective-transient period because the characteristic plume structure exists on initialization. This is illustrated by the $Nu$ spike at $t \sim 15$ in the convectional simulation.}
%     \label{fig:nu_sim}
% \end{figure}
The marginally-stable equilibria are associated with a combination of eigenfunctions whose average kinetic energy is significantly larger than the statistical steady state, as given by Figure \ref{fig:ke_sim}. 
This is postpones kinetic relaxation because the large horizontal flows decay on a viscous timescale
\begin{equation}
    t_{\nu} \sim \sqrt{\frac{Ra}{Pr}}. \nonumber
\end{equation}
Consequently, marginally-stable thermally-equilibrated initial conditions do not reduce the simulation time required to achieve a statistically steady state--rather, they increase it considerably. This suggests the MSTE background state perspective is, at least partially, flawed as an appropriate background state would approximate $\langle|\mathbf{u}|^2 \rangle_{\mathcal{D}}$ with more fidelity.
% \begin{figure}
%     \centering
%     \includegraphics[width=3.4in]{sim_eq_ke.png}
%     \caption{Average kinetic energies of the same simulations performed in Figure \ref{fig:nu_sim} ($Ra = 10^8$). The combination of eigenfunctions which give thermal equilibrium have significantly more kinetic energy than the statistically steady state. Kinetic equilibrium is achieved according to the viscous timescale $t_{\nu} \sim Pr Ra^{-1/2}$.}
%     \label{fig:ke_sim}
% \end{figure}
\begin{figure}
    \begin{minipage}{3.4in}
        \centering
        \includegraphics[width=3.4in]{sim_eq_nu.png}
        \caption{Nusselt numbers of simulations performed at $Ra = 10^8$ with the conventional initial conditions (black) and MSTE initial conditions (blue). 
        MSTE simulations do not undergo a convective-transient period because the characteristic large-scale convective cell structure exists on initialization. For high $Ra$ simulations, researchers often choose to recycle the results of similar $Ra$ runs \cite{Verzicco, Johnston} because the high-Reynolds number flows associated with the convective transient period are computationally prohibitive \cite{Anders_AE}.}
        \label{fig:nu_sim}
    \end{minipage}
\end{figure}

@@

We cannot use our reduced quasilinear model to predict $\Nu \sim Ra^{\beta}$ for large $Ra$ without some complementary stability constraint. 
The eigenvalues of DNS data might provide an informative stability constraint (we could find MUTE and compare them with DNS data), but this would be self-defeating in general. 
Instead we seek to combine reduced models to approximate $\beta$ without DNS. Under the marginal-stability constraint, we found that MSTE exhibit classical Malkus scaling, having marginally-stable boundary layers with $\beta = 1/3$. 
We could instead require that the eigenvalues of MUTE agree with the eigenvalues of the analytic 1-dimensional boundary layer predictions of \cite{Shishkina, Zhang_20} (which remain in development), allowing us to refine and expand these estimates into 2-dimensional MUTE. 
Accessible MUTE states would not rely on DNS, satisfying only the quasilinear problem and stability constraint. 
Such MUTE ought to agree with DNS data more precisely then MSTE, ideally having less extreme velocities. 
This is reasonable to expect, since accurate boundary layer approximations and DNS data should have similar stability characteristics. 
By definition, $\bar{T}_{\rm{MUTE}}$ is unstable and would therefore admit more diffuse boundary layer structures, requiring smaller $A^2(t)$ (less advection) to satisfy the stability constraint and thereby reducing $\Nu$, $\langle |\mathbf{u}|^2 \rangle_{\mathcal{D}}$ and possibly $\beta$.
